The Spatial Model Functions

### 2008-12-15 pkg/diseasemapping/R/New Folder/SpatialFun/SpatialFun.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@18 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Let $Y_k$ denotes the number of cancer cases in group k.
$P_k$ denotes the number of people (population size).
$\lambda_k$ denotes the rate of getting cancer.

We have
     \[Y_k \sim Poisson (P_k\lambda_k),\]  we won't put time under our consideration at the beginning.  %or the function should be Y_k ~ Poisson(\lambda_i T)
     \[\log(\lambda_k) = \mu +  x_k\beta + U_k + V_k,\]
where      $\mu$ is the intercept, it could be $\log(P_k)$;

           $x_k$ is the covariates i.e income variable;

@@

           $U_k$ is the spatial random variable and \[U_k \sim N(0, \sigma^2_u)\]

           $V_k$ is the non-spatial random variable and \[V_k \sim N(0, \sigma^2_v)\]

@@

> data(casedata)
> head(casedata)

@@

> formatCases(casedata)
> head(casedata)

After applying the \texttt{formatCases} function, it will be:
\begin{table}[ht]
\begin{center}
\begin{tabular}{cccccccc}
 Year& $CSD2006$ & $CD2006$ & $PR2006$ & AGE\_SEX\_GROUP &Cases &sex &age    \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $210$   &  $3$ &  F  &$50$ \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $213$   &  $1$ &  F  &$65$ \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $214$   &  $6$ &  F  &$70$ \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $215$   &  $1$ &  F  &$75$ \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $216$   &  $3$ &  F  &$80$ \\
 \end{tabular}
\end{center}
\end{table}

@@

> data(popdata)
> head(popdata@data)

\begin{table}[ht]
\begin{center}
\begin{tabular}{ccccccc}

  CSDUID   &     CSDNAME      & $M0\_4$& $M5\_9$ &\dots & $F0\_4$& $F5\_9$ \\
 $3501005$ & South Glengarry  & $295$ & $355$  &\dots & $285$ & $320$  \\
 $3501011$ & South Stormont   & $310$ & $380$  &\dots & $265$ & $345$  \\
 $3501012$ &      Cornwall    & $1180$& $1295$ &\dots & $1080$& $1210$ \\
 $3501020$ &  South Dundas    & $230$ & $285$  &\dots & $240$ & $300$  \\
 $3501030$ &  North Dundas    & $295$ & $315$  &\dots & $310$ & $285$  \\
 $3501042$ &  North Stormont  & $200$ & $220$  &\dots & $180$ & $220$  \\

@@

After applying the \texttt{formatPopulation} function, it will be:

> formatPopulation(popdata)

\begin{table}[ht]
\begin{center}
\begin{tabular}{cccccccccc}
  CSDUID   & CSDNAME  & $M0\_4$    &\dots &$F0\_4$  & \ldots  &  GROUP &POPULATION& AGE &SEX   \\
 $3501005$ & South Glengarry  & $295$ &\dots & $285$ &\ldots & $M0\_4$& $295$ & $0\_4$ &  M \\
 $3501011$ & South Stormont   & $310$ &\dots & $265$ &\ldots & $M0\_4$& $310$ & $0\_4$ &  M \\ 
 $3501012$ &      Cornwall    & $1180$&\dots & $1080$&\ldots & $M0\_4$& $1180$& $0\_4$ &  M \\ 
 $3501020$ &  South Dundas    & $230$ &\dots & $240$ &\ldots & $M0\_4$& $230$ & $0\_4$ &  M \\ 
 $3501030$ &  North Dundas    & $295$ &\dots & $310$ &\ldots & $M0\_4$& $295$ & $0\_4$ &  M \\ 
 $3501042$ &  North Stormont  & $200$ &\dots & $180$ &\ldots & $M0\_4$& $200$ & $0\_4$ &  M \\ 
\end{tabular}
\end{center}
\end{table}

There are four new columns added in: \texttt{Group}, \texttt{POPULATION}, \texttt{AGE} and \texttt{SEX}.

If we ran the following code:

cancerRates = getRates(casedata, popdata, ~age*sex)

The result of model will be the predicted coefficients of different age and sex groups:

cancerRates

\begin{verbatim}
Call:  glm(formula = formula1, family = family, data = newdata)

Coefficients:
  (Intercept)       age70_74       age60_64       age75_79       age80_84       age55_59       age85_89       age50_54  
     -4.72155        0.18543       -0.38010        0.07425        0.44084       -0.86524        0.30266       -1.86916  
     age45_49       age40_44       age35_39       age30_34       age25_29       age20_24       age15_19         age5_9  
     -1.99423       -2.76184       -3.64470       -5.13532       -4.74062       -6.79155       -7.14367       -6.99895  
     age10_14         age0_4           sexF  age70_74:sexF  age60_64:sexF  age75_79:sexF  age80_84:sexF  age55_59:sexF  
     -7.16576       -8.02353       -0.31172       -0.17597        0.01024       -0.02530       -0.67770             NA  
age85_89:sexF  age50_54:sexF  age45_49:sexF  age40_44:sexF  age35_39:sexF  age30_34:sexF  age25_29:sexF  age20_24:sexF  
     -0.66813        0.42881        0.05375        0.81430        0.72865        1.25281       -0.30888        1.01391  
age15_19:sexF    age5_9:sexF  age10_14:sexF    age0_4:sexF  
           NA             NA             NA             NA

Degrees of Freedom: 30 Total (i.e. Null);  0 Residual
Null Deviance:      38660 
Residual Deviance: 3.535e-13    AIC: 278.9 
\end{verbatim}
The above is the typical result from running the \texttt{glm} model. The $0.18543$, $-0.38010$ etc are the \log of relative risk i.e \beta's.
The base line is %%%%%%%%%

After getting those estimated log relative risk, we can predict the expected number of cancer cases and calculate SMR using function 
\texttt{getSMR}.

ontario = getSMR(cancerRates, popdata, casedata) 
head(ontario@data)

\begin{table}[ht]
\begin{center}
\begin{tabular}{cccccccccc}
  CSDUID   & CSDNAME  & $M0\_4$    &\dots &$F0\_4$  & \ldots  &  GROUP &POPULATION& AGE &SEX   \\
 $3501005$ & South Glengarry  & $295$ &\dots & $285$ &\ldots & $M0\_4$& $295$ & $0\_4$ &  M \\
 $3501011$ & South Stormont   & $310$ &\dots & $265$ &\ldots & $M0\_4$& $310$ & $0\_4$ &  M \\ 
 $3501012$ &      Cornwall    & $1180$&\dots & $1080$&\ldots & $M0\_4$& $1180$& $0\_4$ &  M \\ 
 $3501020$ &  South Dundas    & $230$ &\dots & $240$ &\ldots & $M0\_4$& $230$ & $0\_4$ &  M \\ 
 $3501030$ &  North Dundas    & $295$ &\dots & $310$ &\ldots & $M0\_4$& $295$ & $0\_4$ &  M \\ 
 $3501042$ &  North Stormont  & $200$ &\dots & $180$ &\ldots & $M0\_4$& $200$ & $0\_4$ &  M \\ 
\end{tabular}
\end{center}
\end{table}

>>>>

Let $Y_k$ denotes the number of cancer cases in group k.
$P_k$ denotes the number of people (population size).

We have
     \[Y_k \sim Poisson (P_k\lambda_k),\]  we won't put time under our consideration at the beginning.  %or the function should be Y_k ~ Poisson(\lambda_i T)
     \[\log(\lambda_k) = \mu +  x_k\beta + U_k + V_k,\]
where      $\mu$ is the intercept, it could be \log(P_k);

           $x_k$ is the covariates i.e income variables;

@@

           $U_k$ is the spatial random variable and $U_k$ \sim $N(0, \sigma_u^2)$

           $V_k$ is the non-spatial random variable and $V_k$ \sim $N(0, \sigma_v^2)$

@@

@@

After applying the \texttt{formatCases} function, it will be:
\begin{table}[ht]
\begin{center}
\begin{tabular}{cccccccc}
 Year& $CSD2006$ & $CD2006$ & $PR2006$ & AGE\_SEX\_GROUP &Cases &sex &age    \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $109$   &  $1$ &  M  &$45$ \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $110$   &  $1$ &  M  &$50$ \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $111$   &  $3$ &  M  &$55$ \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $112$   &  $1$ &  M  &$60$ \\
 $1999$& $3501005 $ & $3501 $  &  $35 $     &     $113$   &  $2$ &  M  &$65$ \\
 \end{tabular}
\end{center}
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{cccccc}

  CSDUID   &     CSDNAME  & CSDTYPE & PRUID  & PRNAME & CDUID \\
 $3501005$ & South Glengarry  &    TP  & $35$ &Ontario & $3501$ \\
 $3501011$ & South Stormont   &   TP   & $35$ &Ontario & $3501$  \\
 $3501012$ &      Cornwall    &  CY    &$35$  &Ontario & $3501$  \\
 $3501020$ &  South Dundas    &  TP    &$35$  &Ontario & $3501$  \\
 $3501030$ &  North Dundas    &  TP    &$35$  &Ontario & $3501$  \\
 $3501042$ &  North Stormont  &    TP  &  $35$& Ontario&  $3501$ \\

@@

After applying the \texttt{formatPopulation} function, it will be:
\begin{table}[ht]
\begin{center}
\begin{tabular}{cccccccc}
   CSDUID &    cases&\ldots & CDUID & \ldots &expected    &logExpected  &     SMR    \\
 $3501005$&    $28$& \ldots & $3501$& \ldots &$8.433922$  &  $2.132262$ &$3.319926 $ \\
 $3501007$&    $ 0$& \ldots & $3501$& \ldots &$      NA$  &  $      NA$ &$      NA $ \\
 $3501011$&    $24$& \ldots & $3501$& \ldots &$7.143624$  &  $1.966220$ &$3.359639 $ \\
 $3501012$&    $52$& \ldots & $3501$& \ldots &$28.673675$ &  $ 3.355979$&$ 1.813510$ \\
 $3501020$&    $22$& \ldots & $3501$& \ldots &$6.654331$  &  $1.895268$ &$3.306117 $ \\
 $3501030$&    $19$& \ldots & $3501$& \ldots &$5.855935$  &  $1.767456$ &$3.244572 $ \\
 \end{tabular}
 \end{center}
 \end{table}

There are four new columns added in: cases, this is the number we got from the case data set;
expected, this is the predicted number of cases we calculated using the glm function;
logexpected are the \mbox{log} values of the expected numbers; SMR is the rate of \[\frac {oberved(cases)}{expected(expected)}.\]

if we ran the following code:
\begin{verbatim}
> model <- getRates( casedata, popdata, ~age*sex, S=c("M","F"))
\end{verbatim}

The result of model will be the predicted coefficients of different age and sex groups:
\begin{verbatim}
> model

Call:  glm(formula = formula, family = family, data = newdata)

Coefficients:
(Intercept)         age5        age10        age15        age20        age25
  -13.41885     -0.07403      0.52666      0.87985      1.62710      2.35525
      age30        age35        age40        age45        age50        age55
    3.49147      4.32618      5.12292      5.49001      5.96051      5.79087
      age60        age65        age70        age75        age80         sexM
    7.03365      7.11180      7.22133      7.24461      7.10451      0.67377
  age5:sexM   age10:sexM   age15:sexM   age20:sexM   age25:sexM   age30:sexM
         NA           NA           NA     -1.08827     -0.20732     -1.65308
 age35:sexM   age40:sexM   age45:sexM   age50:sexM   age55:sexM   age60:sexM
   -0.81485     -1.05938     -0.60803     -0.59874           NA     -0.73841
 age65:sexM   age70:sexM   age75:sexM   age80:sexM
   -0.38252     -0.25474     -0.20245           NA

Degrees of Freedom: 28 Total (i.e. Null);  0 Residual
Null Deviance:      10680
Residual Deviance: -5.529e-14   AIC: 229.2

\end{verbatim}
